

\subsection{Solving Mutual Exclusion - Software Level}

This discusses how to solve mutual exclusion from a software point of view.

\subsubsection{Atomic Read/Write Registers}

An atomic shared register satisfies the following properties:
\begin{itemize}
    \item Each invocation $op$ of a \texttt{read} or \texttt{write} operation:
    \begin{itemize}
        \item Appears as if it was executed at a single point $T(op)$ of the timeline.
        \item $T(op)$ is such that $T_{start}(op) \le T(op) \le T_{end}(op)$
        \item For any two operation invocations $\mathnormal{op}_1$ and $\mathnormal{op}_2$ we have that
        $\mathnormal{op}_1 \neq \mathnormal{op}_1 \Rightarrow T(\mathnormal{op}_1) \neq T(\mathnormal{op}_1)$.
    \end{itemize}
    \item Each read invocation returns the value written by the closest preceding write invocation in the sequence defined by the $T(\dots)$ instants associated with the operation invocations.
\end{itemize}

\subsubsection{After You}

We assume there are only two processes trying to access the critical region.

\begin{lstlisting}[
    caption=Acquire Mutex Operation.
]
fn acquire_mutex(i) {
  AFTER_YOU = i;
  wait(AFTER_YOU != i);
}
\end{lstlisting}

\begin{lstlisting}[
    caption=Release Mutex Operation.
]
fn release_mutex(i) { }
\end{lstlisting}

While this approach provides mutual exclusion, it does not guarantee progress.
Furthermore this approach can cause starvation, for example,
if process $p_0$ tries to access the critical region and the other process $p_1$ is busy,
$p_0$ will not be able to access the critical region since the variable \texttt{AFTER\_YOU} will not change.

\subsubsection{Flag Array}

\begin{lstlisting}[
    caption=Acquire Mutex Operation.
]
fn acquire_mutex(i) {
  FLAG[i] = up;
  wait(FLAG[j] == down);
}
\end{lstlisting}

\begin{lstlisting}[
    caption=Release Mutex Operation.
]
fn release_mutex(i) {
  FLAG[i] = down;
}
\end{lstlisting}

This approach provides mutual exclusion, however, once again it does not guarantee progress as it may cause a deadlock\footnote{A \textit{deadlock} happens when two processes are waiting on each other.}.
As we can see in \autoref{eq:interleaving}, if both processes are able to put their flag to \texttt{up} they will be waiting for each other indefinitely.
\begin{equation}\label{eq:interleaving}
    \begin{split}
        p_0 & \rightarrow \mathtt{FLAG[0] = up}\\
        p_1 & \rightarrow \mathtt{FLAG[1] = up}\\
        p_0 & \rightarrow \mathtt{wait}\\
        p_1 & \rightarrow \mathtt{wait}\\
    \end{split}
\end{equation}

\subsubsection{Flag Array with Delayed Wait}

\begin{lstlisting}[
    caption=Acquire Mutex Operation.
]
fn acquire_mutex(i) {
  FLAG[i] = up;
  while (FLAG[j] == up) {
    FLAG[i] = down:
    sleep(random());
    FLAG[i] = up;
  }
}
\end{lstlisting}

\begin{lstlisting}[
    caption=Release Mutex Operation.
]
fn release_mutex(i) {
  FLAG[i] = down;
}
\end{lstlisting}

This solution provides mutual exclusion but is not able to guarantee progress as well.
However only in specific cases where the processes are synchronized and the result from the \texttt{random()} call is consistently the same for both processes the program will be stuck,
creating a livelock\footnote{A livelock is similar to a deadlock, except that the states of the processes involved in the livelock constantly change with regard to one another, none progressing.}.

This solution may also suffer from starvation,
in the case that a process $p_0$ acquires the mutex and $p_1$ is waiting,
it is possible for $p_0$ to release and re-acquire the mutex without $p_1$ getting access.

\subsubsection{Mixed Approach}\label{sec:mixed}

Just like the previous schemes,
this one is also applicable for two processes only.


\begin{lstlisting}[
    caption=Acquire Mutex Operation.
]
fn acquire_mutex(i) {
  FLAG[i] = up;
  AFTER_YOU = i;
  wait(
    FLAG[j] == down ||
    AFTER_YOU != i
  );
}
\end{lstlisting}

\begin{lstlisting}[
    caption=Release Mutex Operation.
]
fn release_mutex(i) {
  FLAG[i] = down;
}
\end{lstlisting}

By mixing the previous approaches we are able to achieve progress,
since even if both flags are \texttt{up} the \texttt{AFTER\_YOU} variable will make one proceed to the critical region.
While this solution provides mutual exclusion and progress,
it will only work for two processes.

\subsubsection{Mixed Approach for N processes}

\begin{lstlisting}[
    caption=Acquire Mutex Operation.
]
fn acquire_mutex(i) {
  for idx in 1 to (n-1) {
    FLAG_LEVEL[i] = idx;
    AFTER_YOU[idx] = i;
    wait(
      forall k != i :
      FLAG_LEVEL[k] < idx ||
      AFTER_YOU[idx] != i
    );
  }
}
\end{lstlisting}

\begin{lstlisting}[
    caption=Release Mutex Operation.
]
fn release_mutex(i) {
  FLAG_LEVEL[i] = 0;
}
\end{lstlisting}

This approach generalizes the one described in \autoref{sec:mixed},
allowing for $N$ processes to use the mutex.
The intuition behind it so that whenever $n$ processes try to access the critical region,
they will go through $n-1$ levels, the process reaching the last one will gain access to the region.

$p_i$ is allowed to progress to level $idx + 1$ if, from its point of view, one of the following is true:
\begin{itemize}
    \item All other processes are at a lower level ($\forall k \neq i : \mathtt{FLAG\_LEVEL[k] < idx}$).
    \item It was not the last one to enter the level $idx$ ($\mathtt{AFTER\_YOU[idx]} \neq i$).
\end{itemize}
